---
title: "Lab 10: Halloween Mini-Project"
author: "Caroline Mackey (A15522472)"
date: "10/28/2021"
output: pdf_document
---

## Importing candy data. 

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy <- read.csv(candy_file, row.names=1)
head(candy)
```

> **Q1.** How many different candy types are in this dataset?

```{r}
nrow(candy)
```

**A1.** There are 85 different candy types in this data set. 


> **Q2.** How many fruity candy types are in the dataset?


```{r}
sum(candy$fruity)
```

**A2.** There are 28 fruity candy types in the dataset.

## 2. What is your favorite candy? 

```{r}
candy["Twix",]$winpercent
```

> **Q3.** What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy in the data set is Peanut butter M&M's. 

```{r}
candy["Peanut butter M&MÕs",]$winpercent
```

It's win percent is approximately 71.5%. 

> **Q4.** What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

The win percent value for Kit Kat is approximately 76.8%. 

> **Q5.** What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

The win percent value for Tootsie Roll Snack Bars is approximately 50.0%. 

#### Skimr
Download the skimr package to give overview of dataset. 

```{r}
library("skimr")
skim(candy)
```

> **Q6.** Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

**A6.** Yes, most of the rows have values ranging from 0 to 1, while winpercent ranges from 0 to 100. 

> **Q7.** What do you think a zero and one represent for the candy$chocolate column?

**A7.** In the candy$chocolate column, a zero represents that the candy **is not** chocolate-y, while a one represents that the candy **is** chocolate-y. 

> **Q8.** Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot (candy, aes(x=candy$winpercent)) + 
  geom_bar()
```


> **Q9.** Is the distribution of winpercent values symmetrical?

**A9.** No. 

> **Q10.** Is the center of the distribution above or below 50%?

**A10.** Below. 

> **Q11.** On average is chocolate candy higher or lower ranked than fruit candy?

First need to find all the chocolate candy row in the 'candy' dataset. 

```{r}
chocolate <- as.logical(candy$chocolate) 
chocolate_win <- candy[chocolate,]$winpercent
chocolate_win
```

Do the same for fruity candy: 

```{r}
fruity <- as.logical(candy$fruity)
fruity_win <- candy[fruity,]$winpercent
fruity_win
```

Compare chocolate vs. fruity candy. 

```{r}

# Average chocolate win percent... 
chocolate_win_avg <- mean(chocolate_win)
chocolate_win_avg

# Average fruity  win percent... 
fruity_win_avg <- mean(fruity_win)
fruity_win_avg

# Is average chocolate win percent greater than the average fruity win percent? 
chocolate_win_avg > fruity_win_avg

```

**A11.** Choclate candy is higher ranked than fruity candy. 

> **Q12.** Is this difference statistically significant?

T-test!

```{r}
t.test(chocolate_win,fruity_win)
```

**A12.** A t-test produces a p-value of **2.871e-0.8**, which is much smaller than the typical threshold of  0.05. Yes, it is statistically significant. 

## 3. Overall Candy Rankings 

> **Q13.** What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```
**A13.** Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters are the **least** liked candy types (i.e. the candy types with the lowest winpercent). 

> **Q14.** What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(-candy$winpercent),], n=5)
```
**A14.** Reese's Peanut Butter Cups, Reese's Miniatures, Twix, Kit Kats, and Snickers are the **most** liked candy types (i.e. the candy types with the highest winpercent). 

> **Q15.** Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(data=candy) + 
  aes(winpercent, rownames(candy)) + 
  geom_col()
```

> **Q16.** This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(data=candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col()
```

#### Add some useful color... 

Make a color vector. 

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate4"
my_cols[as.logical(candy$bar)] = "darkorchid4"
my_cols[as.logical(candy$fruity)] = "darkorange"
```

Add color vector to the plot

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) + 
  labs (title="Win Percent of Various Halloween Candies",
        x=("Win Percent"), y=("Halloween Candy"))
```

> **Q17.** What is the worst ranked chocolate candy?


> **Q18.** What is the best ranked fruity candy?


## 4. Taking a look at pricepercent

```{r}
# How about a plot of price vs win
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 10) + 
  labs (title="Price Percent vs. Win Percent of Various Halloween Candies",
        x=("Win Percent"), y=("Price Percent"))
```

> **Q19.** Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

> **Q20.**  What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

## 5. Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

## 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Now we can plot our main PCA score plot of PC1 vs PC2.
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> **Q24.** What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

**A24.** Positive direction:fruity, hard, pluribus; negative direction: chocolate, caramel, peanuty-almondy, nougat, crisper-rice-wafer, bar, sugar percent, price percent, win percent 