---
title: "Lab 17: Vaccination rate mini project"
author: "Caroline Mackey (A15522472)"
date: "11/23/2021"
output: 
  html_document:
    latex_engine: xelatex
---

## Background

We are investigating data on vaccination rates in California.

## Getting started

```{r}
# Import vaccination data 

vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax) 
```

> **Q1.** What column details the total number of people fully vaccinated?
>
> **A1.** The column **persons_fully_vaccinated** details the total number of people fully vaccinated.

> **Q2.** What column details the Zip code tabulation area?
>
> **A2.** The column **zip_code_tabulation_area** details the Zip code tabulation area.

> **Q3.** What is the earliest date in this data set?

```{r}
vax$as_of_date[1]
```

> **A3.** **2021-01-05** is the earliest date in the data set.

> **Q4.** What is the latest date in this data set?

```{r}
vax$as_of_date[nrow(vax)]
```

> **A4.** **2021-11-16** is the latest date in the data set.

Use the **skim()** function to quickly overview & summarize the various columns of the data set.

```{r}
skimr::skim(vax)
```

> **Q5.** How many numeric columns are in this dataset?

```{r}
library("dplyr")
length(select_if(vax,is.numeric))
```

> **A5.** There are **9** numeric columns in the vax data set.

> **Q6.** Note that there are "missing values" in the data set. How many NA values there in the persons_fully_vaccinated column?

```{r}
sum.na <- sum( is.na(vax$persons_fully_vaccinated) ) 
sum.na
```

Another method:

```{r}
summary(vax$persons_fully_vaccinated)
```

> **A6.** There are **8256** missing values (i.e. "NA"s) in the persons_fully_vaccinated column.

> **Q7.** What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
sum.na/nrow(vax)
```

> **A7.** Approximately **10.%** of persons_fully_vaccinated values are missing.

> **Q8.** \[Optional\]: Why might this data be missing?

> **A8.** One reason for the missing data could have been if there were issues reporting/ obtaining the data due to confidentiality.

## Working with dates

Use the **lubridate** package to make life a lot easier when dealing with dates & times.

```{r}
library(lubridate)
```

Make our `as_of_date` column lubridate format.

```{r}
# Specify that we are using the year-month-day format. 
vax$as_of_date <- ymd(vax$as_of_date)
```

Now, we can do useful math with dates more easily.

```{r}
# Example 
today() - vax$as_of_date[1]
```

> **Original Q9.** How many days have passed since the last update of the data set?

```{r}
today()
today()- vax$as_of_date[nrow(vax)]
```

> **Original A9.** 7 days have passed since the last update of the data set and today (November 23).

> **New Q9.** How many days between the first & last enry in the dataset? (changed this question in class)

```{r}
# Last entry: vax$as_of_date[nrow(vax)]
# First entry: vax$as_of_date[1]
# Take the difference
vax$as_of_date[nrow(vax)]-vax$as_of_date[1]
```

> **New A9.** **315 days** have passed between the first & last entry of the data set.

> **Q10.** How many unique dates are in the data set (i.e. how many different dates are detailed)?

```{r}
# unique(vax$as_of_date) prints all of the unique dates 
# take the length of this to print how many 
length(unique(vax$as_of_date))
```

> **A10** There are **46** unique dates in the data set.

## Working with ZIP Codes

We will use the **zipcodeR** package to help make sense of the ZIP Code data.

```{r}
library(zipcodeR)
```

La Jolla ZIP Code:

```{r}
geocode_zip('92037')
```

Distance between 2 ZIP Codes:

```{r}
zip_distance('92037','92109')
```

Census data:

```{r}
reverse_zipcode(c('92037', "92109") )
```

```{r}
# Pull data for all ZIP codes in the dataset
# zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

## Focus on the San Diego area

We will subset with base R

```{r}
# Subset to San Diego county only areas
sd <- vax$county == "San Diego"
head(vax[sd,])
```

```{r}
library(dplyr)
sd <- filter(vax, county=="San Diego")
nrow(sd)
```

Subsetting all San Diego county areas with a population of over 10,000

```{r}
sd.10 <- filter(vax, county == "San Diego" & 
                  age5_plus_population > 1000)
```

> **Q11.** How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

> **A11.** There are **107** distinct ZIP codes for San Diego County.

> **Q12.** What San Diego County Zip code area has the largest 12 + Population in this data set?

```{r}
max12pop <- which.max(sd$age12_plus_population)
sd[max12pop,]
```

> **A12.** ZIP Code **92154** has the largest 12+ population in this data set.

> **Q13.** What is the overall average "Percent of Population Fully Vaccinated" value for all San Diego "County" as of "2021-11-09"?

```{r}
sd.11.9 <- filter(sd, as_of_date=="2021-11-09")
head(sd.11.9)
mean(sd.11.9$percent_of_population_fully_vaccinated, na.rm=TRUE)
```

> **A13.** The average Percent of Population Fully Vaccinated value for all San Diego County as of 2021-11-09 is **67.3%** .

We could also look at the 6-number summary.

```{r}
summary(sd.11.9$percent_of_population_fully_vaccinated)
```

> **A14.** Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of "2021-11-09"?

```{r}
library(ggplot2)

ggplot(sd.11.9) + 
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins=15) + 
  labs (x="Percent of Population Fully Vaccinated in a ZIP Code Area",
        y="Count (ZIP Code areas", 
        title="Histogram of Vaccination Rates Accross San Diego County",
       subtitle="As of 2021-11)-09") 
```

> \*A14.\*\* See graph above.

## Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area == "92037")
ucsd
```

```{r}
# Example: what is the age 5+ population in the 92037 ZIP code? 
ucsd[1,]$age5_plus_population
```

> **Q15.** Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) + 
  aes(x=as_of_date, 
      y=percent_of_population_fully_vaccinated) + 
  geom_point() + 
  geom_line(groups=1) + 
  ylim(c(0,1)) + 
  labs(x="Date", y="Percent Vaccinated")
```

> **A15.** See graph above.

## Comparing 92037 to other similar sized areas?

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

#head(vax.36)
```

> **Q16.** Calculate the mean "Percent of Population Fully Vaccinated" for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date "2021-11-16". Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean(vax.36$percent_of_population_fully_vaccinated)
```

```{r}
ggplot(ucsd) + 
  aes(x=as_of_date, 
      y=percent_of_population_fully_vaccinated) + 
  geom_point() + 
  geom_line(groups=1) + 
  ylim(c(0,1)) + 
  labs(x="Date", y="Percent Vaccinated") + 
  geom_hline(yintercept = 0.66, linetype="longdash", color="red")
```

> **A16.** The mean Percent of Population Fully Vaccinated for ZIP code areas with a population as large as 92037 (La Jolla) as of 2021-11-16 is approximately **66%**. This value is addded to the graph above as a red line.

> **Q17.** What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the "Percent of Population Fully Vaccinated" values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date "2021-11-16"?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> **A17.** See summary above.

> **Q18.** Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) + 
  aes (x=percent_of_population_fully_vaccinated) + 
  geom_histogram() + 
   labs(x="Percent Vaccinated", y="Count")
```

> **A18** See graph above.

> **Q19.** Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
# Print mean again 
mean(vax.36$percent_of_population_fully_vaccinated)
```

```{r}

filter(vax.36, zip_code_tabulation_area=="92109")
```

```{r}
filter(vax.36, zip_code_tabulation_area=="92040")
```

> **A19.** The percent of population fully vaccinated in the **92109** ZIP code is approximately **69%**, which is higher than average of approximately 66%. The percent of population fully vaccinated in the **92040** ZIP code is approxumately **52%**, which is lower than average.

```{r}
colnames(vax.36)
```

> **Q20.** Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population \> 36144

First, we need to subset the full `vax` daraset to include only ZIP codes areas with a population as large as 92037.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
nrow(vax.36.all)
```

How many unique zip codes have a population as large as 92307?

```{r}
length(unique(vax.36.all))
```

Let's make a final figure that shows all these ZIP areas.

```{r}
ggplot(vax.36.all) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") + 
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination rate across California",
       subtitle="Only areas with a population above 36k are shown") + 
  geom_hline(yintercept = 0.66, linetype="longdash")
```

> **A20.** See graph above.

> **Q21.** How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

> **A21.** I still feel comfortable travelling for Thanksgiving and meeting for in-person class, especially since I am vaccinated and I now know a lot of people in the areas I will be are also vaccinated. However, it's always a priority to keep those around me safe, so I still tend to stay on the safe side of things, like by avoiding large crowds.

Have a great break! :)
