---
title: "Lab 9 Mini-Project (NEW)"
author: "Caroline Mackey (A15522472)"
date: "11/1/2021"
output: html_document
---

## 1. Exploratory data analysis

#### Preparing the data 

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)

# Check that column names are correct 
head(wisc.df)
```

Remove the first column, since that is the “answer” to our problem…

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
head(wisc.data)
```

```{r}
# Create diagnosis vector for later 
diagnosis <- factor(wisc.df$diagnosis) 
```

#### Exploratory data analysis

> **Q1.** How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

> **A1.** There are 569 observations in this data set.

> **Q2.** How many of the observations have a malignant diagnosis?

```{r}
table(diagnosis)
```

> **A2.** There are 212 malignant diagnoses.

> **Q3.** How many variables/features in the data are suffixed with _mean?

```{r}
names <- colnames(wisc.df)
mean <- grep("_mean", names)
length(mean)
```

> **A3.** 10 variables or features in the data are suffixed with _mean.

## 2. Principal Component Analysis

First, see if the data needs to be scaled.

```{r}
# Check column means and standard deviations
colMeans (wisc.data)
apply(wisc.data,2,sd)
```

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale=TRUE)

# Look at summary of results. 
summary(wisc.pr)
```

> **Q4.** From your results, what proportion of the original variance is captured by the first principal components (PC1)?

> **A4.** 44.27% of variance is captuerd by PC1.


> **Q5.** How many principal components (PCs) are required to describe at least 70% of the original variance in the data?


> **A5.** 3 PCs are required to describe at least 70% of the original variance in the data.


> **Q6.**  How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

> **A6.** 7 PCs are required to describe at least 80% of the original variance in the data.

#### Interpreting PCA results

Create a biplot. 

```{r}
biplot (wisc.pr)
```
> **Q7.** What stands out to you about this plot? Is it easy or difficult to understand? Why?

> **A7.** Nothing significant stands out from this plot, as it is very difficult to understand. We will need to make our own!

Generate a new, standard scatter-plot. 

```{r}
# Scatter plot observations by components 1 and 2
plot(x=wisc.pr$x[,1], y=wisc.pr$x[,2], 
     col = diagnosis , 
     xlab = "PC1", ylab = "PC2")
```

> **Q8**  Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(x=wisc.pr$x[,1], y=wisc.pr$x[,3], 
     col = diagnosis , 
     xlab = "PC1", ylab = "PC3")
```

> **A8.** The two graphs look relatively similar. They both have two color-coded clusters of malignant vs. benign cells. It’s clear that there is something that delineates the two types of cells, so there are some characteristics that define benign cells.

Using ggplot. 

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

#### Variance explained

Produce scree plots to see proportion of variance.

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

Calculate the variance explained by each principal component by dividing by the total variance explained of all principal components.

```{r}
# Variance explained by each principal component: pve
pve <- pr.var/sum(pr.var) 

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

#### Communicating PCA results

> **Q9.** For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```

> **A9.** For the first principal component, the component of the loading vector for the feature concave.points_mean is -0.2608538. 

> **Q10.** What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
summary(wisc.pr)
```

> **A10.** A minimum of 5 PCs is required to explain 80% of the variance of the data?

## 3. Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
# Calculate the (Euclidean) distances between all pairs of observations in the new scaled dataset
data.dist <- dist(data.scaled)
# Create a hierarchical clustering model using complete linkage.
wisc.hclust <- hclust(data.dist)
```

#### Results of hierarchical clustering

> **Q11.** Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```

> **A11.** At a height of approximately 19, the clustering model has 4 clusters (see above).

#### Selecting number of cluster

Use cutree() to cut the tree so that it has 4 clusters. Assign the output to the variable wisc.hclust.clusters.

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters, diagnosis)
```

> **Q12.** Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
wisc.hclust.clusters2 <- cutree(wisc.hclust, k=2)
table(wisc.hclust.clusters, diagnosis)
```

> **A12** k=2 seems to be a better cluster vs. diagnoses match. This would be the expected number of beningn vs. malignant cells. 


> **Q13.** Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
hc.complete <- hclust (data.dist, method="complete") 
plot(hc.complete)

hc.average <- hclust (data.dist, method="average")
plot(hc.average)

hc.single <- hclust (data.dist, method="single") 
plot(hc.single)

hc.ward.D2 <- hclust (data.dist, method="ward.D2") 
plot(hc.ward.D2)
```

> **A13** My favorite method is also ward.D2 since the branches seem to be the most evenly distributed and it's easiest to read. 


## 4. OPTIONAL: K-means clustering

```{r}
wisc.km <- kmeans(scale(wisc.data), centers= 2, nstart= 20)
table(wisc.km$cluster, diagnosis)
```

> **Q14.** How well does k-means separate the two diagnoses? How does it compare to your hclust results?

> **A14.** The k-means seems to separate the two diagnoses well, with the first group being mostly benign cells, and the second grou being mostly malignant cells. Group 1 in this data set seems to align with group 3 in the hclust results, while Group 2 in this data set seems to align with group 1 in the hclust restuls. 

```{r}
table(wisc.hclust.clusters, (wisc.km$cluster))
```

## 5. Combining methods

#### Clustering on PCA results

```{r}
wisc.pr.hclust <- hclust (dist(wisc.pr$x[,1:7]), method="ward.D2") 
plot(wisc.pr.hclust )
```

See if two main branches divide malignant vs. beningn. 

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
# Make cluster 2 (B) black & cluster 1 (M) red. 
plot(wisc.pr$x[,1:2], col=diagnosis)
```


```{r}
g <- as.factor(grps)
levels(g)
g <- relevel(g,2)
levels(g)
# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")
# Cut this hierarchical clustering model into 2 clusters 
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

Using table(), compare the results from your new hierarchical clustering model with the actual diagnoses.

> **Q15.** How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
# Compare to actual diagnoses
table(wisc.pr.hclust.clusters, diagnosis)
```

> **Q15.** The newly created model seems to separate the two diagnoses fairly well. Group 1 is mostly malignant cells, while Group 2 is mostly benign cells. 

> **Q16.** How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.km$cluster, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```

> **A16.** The wisc.hclust.clusters table seems to be pretty similar to the diagnosis data, where Group 1 of wisc.hclust.clusters aligns with the malignant diagnosis, and Group 3 of wisc.hclust.clusters aligns with the benign diagnosis. 

## 6. Sensitivity/Specificity

> **Q17**  Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

Sensitivity: TP/(TP+FN)

```{r}
# wisc.km$cluster
(175)/(175+14)

# wisc.pr.hclust.clusters
(188)/(188+28)
```

Specificity: TN/(TN+FN)

```{r}
# wisc.km$cluster
(343)/(343+37)

# wisc.pr.hclust.clusters
(329)/(329+24)
```

> **A17.** The k-means approach is better sensitivity while the hierarchical clustering approach has better specificity. 

## 7. Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> **Q18.** Which of these new patients should we prioritize for follow up based on your results?

> **A18.** We should prioritize Patient 2, since their datapoint falls in the red cluster of malignant cells. 