---
title: "Class 6: R Functions"
author: "Caroline Mackey, PID: A15522472"
date: "10/14/2021"
output: pdf_document
---

## R functions 

In today's class, we are going to write a function together that grades some student work. 

Questions for today:

> **Q1.** Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}

# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

```

Let's start with student1 and find their average score 
```{r}
mean(student1)
```

But we want to drop the lowest score. 
We could try the **min()** function. 
```{r}
min(student1)
```

The **which.min()** function looks useful
```{r}
which.min(student1)
```

This gives us the position of the lowest score. 
```{r}
# This would be the lowest score 
student1[which.min(student1)]
```

To drop this value I can use minus
```{r}
student1[-which.min(student1)]
```

Now use mean like before. 
```{r}
mean(student1[-which.min(student1)])
```

Try for student 2 (unsuccessful.)
```{r}
mean(student2[-which.min(student2)])
```

We need to remove the NA elements of the vector.
```{r}
mean(student2[-which.min(student2)], na.rm=TRUE)
```
Unsuccessful: this is dropping the 80 instead of the NA (NA is excluded altogether.)

```{r}

```


One new idea/approach is we could repalce the NA (missing homeworks) with 0. 
```{r}
is.na(student2)
```

The element **is.na()** functin returns a logical vector where TRUE elements represent where the NA values are 
```{r}
which(is.na(student2))
```

Now let's make the NA values = 0 
```{r}
student.prime <-  student2
student.prime
student.prime [which(is.na(student2))] = 0
student.prime
```
Put it all together to get the average score dropping the lowest, where we map NA values to 0. 
```{r}
student.prime <-  student2
student.prime [which(is.na(student2))] = 0
mean(student.prime[-which.min(student.prime)])
```

Check work
```{r}
student2
```
```{r}
mean (c(100,90,90,90,90,97,80))
```
Yay!! 


Check for student 3
```{r}
student.3prime <-  student3
student.3prime [which(is.na(student3))] = 0
mean(student.3prime[-which.min(student.3prime)])
```
```{r}
student3
```
```{r}
mean(c(90,0,0,0,0,0,0))
```
It works! We got our working snippet. 

Let's simplify. 
```{r}
x <-  student3
# Map NA values to 0: 
x [which(is.na(student3))] = 0
#Find the mean without the lowest value: 
mean(x[-which.min(x)])
```

Now we can use this as the base of my function
```{r}
grade <- function(x) {
  #Make sure our scores are all numbers 
  x <- as.numeric(x)
  # Map NA values to 0: 
  x [which(is.na(x))] = 0
  #Find the mean without the lowest value: 
  mean(x[-which.min(x)])
}
```
Try student 1: 
```{r}
grade(student1)
```
Student 2: 
```{r}
grade(student2)
```
Student 3: 
```{r}
grade(student3)
```

Now read the full grade profile. 
```{r}
scores <- read.csv("https://tinyurl.com/gradeinput",row.names = 1)
scores
```

Use **apply()** function to apply grades to every student in the csv file of scores... 
```{r}
gradebook = apply (scores, 1, grade)
gradebook
# write 1 because that's what we named the rows 
```

> **Q2.** Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts] 

```{r}
which.max(gradebook)
```
Student #18 is the top scoring student overall in the gradebook. 

> **Q3.**  From your analysis of the gradebook, which homework was toughest on students (i.e. ob tained the lowest scores overall? [2pts] 

We can use the **aplply()** function over the columns by setting the margin=2 argument 
```{r}
hw = apply(scores, 2, mean, na.rm=TRUE)
print(hw)
which.min(hw)
```
On average, Homework #3 had the lowest score, so it was likely the toughest on students. 

> **Q4.** Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

> **Q5.** Make sure you save your Rmarkdown document and can click the “Knit” button to generate
a PDF foramt report without errors. Finally, submit your PDF to gradescope. [1pt]

> **Extension- Class 7**

## Add an extra argument to our function. 

We want to optimally turn the dropping of lowest score off for gradding. 

```{r}
 
grade2 <- function(x, drop.lowest=TRUE) {
 
  # Make sure our scores are all numbers 
  x <- as.numeric(x)
  
  # Map NA values to 0: 
  x [which(is.na(x))] = 0
  
  if (drop.lowest) {
    # Find the mean without the lowest value: 
    mean(x[-which.min(x)])
  # Find the mean with the lowest value: 
  } else {
    mean(x)}
}
```

```{r}
student1
grade2(student1, drop.lowest=TRUE)
grade2(student1, drop.lowest=FALSE)
```


